Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels

### 2017-11-23 articles/formalism/sections/real_space_formalism.tex
reintroduced the discussion on Euler angles in the formalism draft, with significant edits to the discussion given in the old draft.

Next we define the operators which govern the transformations between different representations of the polarization field as follows,
\bar T &=& \qutox_{2 \rm N_{\rm pix} \times 2 \rm N_{\rm pix}} ~~;~~ \bar T^{-1} = \frac{1}{2} \bar T^{\dagger} \,, \\
\tilde T &=& -\qutox_{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms}} ~~;~~ \tilde T^{-1} = \frac{1}{2} \tilde T^{\dagger} \,,
where we have chosen the sign conventions so as to match those used in Healpix.
Using the data vectors and the matrix operators defined above we can now express, in compact notation, the forward and inverse relations between different representations of the polarization data vectors as follows,
\begin{subequations} \label{eq:pol_data_relns}
\bar{X} &=& \bar T * \bar{P} ~~;~~\bar{P} = \frac{1}{2} \bar T^{\dagger} * \bar{X} \,, \\
\bar X &=&  {{}_2\mathcal{Y}} * \tilde X  ~~;~~ \tilde X ={{}_2\mathcal{Y}}^{\dagger} * \bar X  \,, \\
\tilde{X} &=& \tilde T * \tilde{S} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\ 
\bar S &=&  {{}_0\mathcal{Y}} * \tilde S ~~;~~  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} * \bar S \,.
%\tilde X &=&  {{}_2\mathcal{Y}}^{\dagger} * \bar X ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{S} &=& {{}_0\mathcal{Y}}*\tilde S ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\
%\bar{X} &=& \bar T * \bar{P} ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{P} &=& \frac{1}{2} \bar T^{\dagger} * \bar{X} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,. \\
Next we introduce the harmonic space operators, which project the harmonic space data vector to E or B subspace,
\begin{subequations} \label{eq:har_eb_op}
\tilde O_E &=& \bmat \mathbb{1} & \mathbb{0} \\ \mathbb{0} & \mathbb{0} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} }   ~~;~~ \tilde S_E = \tilde O_E* \tilde S \,,\\
\tilde O_B &=& \bmat \mathbb{0} & \mathbb{0} \\ \mathbb{0} & \mathbb{1} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} } ~~; ~~ \tilde S_B = \tilde O_B *\tilde S \,
Note that these harmonic space matrices are idempotent, orthogonal to each other and their sum is an identity matrix as can be explicitly seen via the following relations, 
\begin{subequations} \label{eq:har_op_prop}
\tilde O_E * \tilde O_E&=& \tilde O_E ~~;~~  \tilde O_B * \tilde O_B= \tilde O_B \,,\\
 \tilde O_E * \tilde O_B&=& \mathbb{0} \,, \label{eq:op_eb_ortho}\\ 
 \tilde O_E + \tilde O_B&=& \mathbb{1} \,.
Note that the above relations for these harmonic space operators are exactly valid.  In the following sections we aim to derive the real space analogues of these harmonic space operators.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Real space operators} \label{sec:real_space_operators}
The vector-matrix notation introduced in  \sec{sec:mat_pol_intro} allows for concise book keeping of all the operations involved in the analysis of CMB polarization. In this section we use this notation to derive the real space operators which translate the Stokes vector \vp{}  to the vector of scalars \vs  and vice versa. This vector-matrix notation also allows us to simply derive real space operators for direct decomposition of the Stokes vector \vp{} in to a vector \vp{ E} that correspond to $E$-modes and another vector \vp{B} that corresponds to the $B$-modes of polarization, such that \vp{} = \vp{E} + \vp{B}, without ever evaluating the $E$ \& $B$ fields or their spherical harmonics. 

>>>>

Next we define the operators which govern the transformations between different representations of the polarization field as follows,
%
\beqrys
\bar T &=& \qutox_{2 \rm N_{\rm pix} \times 2 \rm N_{\rm pix}} ~~;~~ \bar T^{-1} = \frac{1}{2} \bar T^{\dagger} \,, \\
\tilde T &=& -\qutox_{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms}} ~~;~~ \tilde T^{-1} = \frac{1}{2} \tilde T^{\dagger} \,,
\eeqrys
%
where we have chosen the sign conventions so as to match those used in Healpix.
Using the data vectors and the matrix operators defined above we can now express, in compact notation, the forward and inverse relations between different representations of the polarization data vectors as follows,
%
\begin{subequations} \label{eq:pol_data_relns}
\beqry 
\bar{X} &=& \bar T * \bar{P} ~~;~~\bar{P} = \frac{1}{2} \bar T^{\dagger} * \bar{X} \,, \\
\bar X &=&  {{}_2\mathcal{Y}} * \tilde X  ~~;~~ \tilde X ={{}_2\mathcal{Y}}^{\dagger} * \bar X  \,, \\
\tilde{X} &=& \tilde T * \tilde{S} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\ 
\bar S &=&  {{}_0\mathcal{Y}} * \tilde S ~~;~~  \tilde S =  {{}_0\mathcal{Y}}^{\dagger} * \bar S \,.
%\tilde X &=&  {{}_2\mathcal{Y}}^{\dagger} * \bar X ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{S} &=& {{}_0\mathcal{Y}}*\tilde S ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,.\\
%\bar{X} &=& \bar T * \bar{P} ~~;~~ \tilde{X} = \tilde T * \tilde{S} \,, \\
%\bar{P} &=& \frac{1}{2} \bar T^{\dagger} * \bar{X} ~~;~~ \tilde{S} = \frac{1}{2}\tilde T^{\dagger} * \tilde{X} \,. \\
\eeqry
\end{subequations}
%
Next we introduce the harmonic space operators, which project the harmonic space data vector to E or B subspace,
%
\begin{subequations} \label{eq:har_eb_op}
\beqry
\tilde O_E &=& \bmat \mathbb{1} & \mathbb{0} \\ \mathbb{0} & \mathbb{0} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} }   ~~;~~ \tilde S_E = \tilde O_E* \tilde S \,,\\
\tilde O_B &=& \bmat \mathbb{0} & \mathbb{0} \\ \mathbb{0} & \mathbb{1} \emat _{2 \rm N_{\rm alms} \times 2 \rm N_{\rm alms} } ~~; ~~ \tilde S_B = \tilde O_B *\tilde S \,
\eeqry
\end{subequations}
%
Note that these harmonic space matrices are idempotent, orthogonal to each other and their sum is an identity matrix as can be explicitly seen via the following relations, 
%
\begin{subequations} \label{eq:har_op_prop}
\beqry
\tilde O_E * \tilde O_E&=& \tilde O_E ~~;~~  \tilde O_B * \tilde O_B= \tilde O_B \,,\\
 \tilde O_E * \tilde O_B&=& \mathbb{0} \,, \label{eq:op_eb_ortho}\\ 
 \tilde O_E + \tilde O_B&=& \mathbb{1} \,.
\eeqry
\end{subequations}
%
Note that the above relations for these harmonic space operators are exactly valid.  In the following sections we aim to derive the real space analogues of these harmonic space operators.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Real space operators} \label{sec:real_space_operators}
The vector-matrix notation introduced in  \sec{sec:mat_pol_intro} allows for concise book keeping of all the operations involved in the analysis of CMB polarization. In this section we use this notation to derive the real space operators which translate the Stokes vector \vp{}  to the vector of scalars \vs  and vice versa. This vector-matrix notation also allows us to simply derive real space operators for direct decomposition of the Stokes vector \vp{} in to a vector \vp{ E} that correspond to $E$-modes and another vector \vp{B} that corresponds to the $B$-modes of polarization, such that \vp{} = \vp{E} + \vp{B}, without ever evaluating the $E$ \& $B$ fields or their spherical harmonics. 
All the real space operators we derive are most conveniently expressed as functions of the Euler angles  $\alpha, ~\beta~\&~ \gamma$ on the sphere. Owing to this, we begin with a brief discussion on Euler angles and present visual method to think about them.

\textit{Euler angles:}  We define the local cartesian coordinate system at any point on the sphere such that the z-axis points along the radial direction, the x-axis is along the vector tangent to the local longitude pointing south and the y-axis is a vector tangent to the local latitude pointing east. The Euler angles $\alpha ,\beta ~\&~ \gamma$ define rotation operations that transforms the local cartesian coordinate system defined at the location $\hat{n}_0 \equiv (\theta_0,\phi_0)$ such that it aligns with the local cartesian coordinate system at the location $\hat{n}_i \equiv (\theta_i,\phi_i)$ \cite{varshalovich}.

Specifically $\alpha$ defines the rotation about the z-axis, $\beta$ defines the rotation about the new y-axis (y1-axis) after the previous rotation and $\gamma$ defines the rotation about the final z-axis (z2-axis) after carrying out the previous two rotations. In more physical terms, these angles can be understood as follow: The rotation by $\alpha$ about the z-axis is such that it aligns the x-axis of the cartesian system at location $\hat{n}_0$ along the great circle in the direction of $\hat{n}_i$.  The rotation by $\beta$ about the y2-axis parallel transports the local cartesian coordinate system from location $\hat{n}_0$ to location $\hat{n}_i$, such that the z-axes of the two coordinate systems are parallel to each other. Finally the rotation by angle $\gamma$ about the z2-axis aligns the x \& y axes of the parallel transported system with those of the local cartesian system defined at $\hat{n}_i$.

The simplest case is when one of the coordinates coincides with the north pole (more formally this refers to the point $\theta_0 \rightarrow 0$ while moving along the longitude $\phi_0=0$), say $\hat{n}_0=(\theta_0,\phi_0)=(0,0)$. In this case it is easy to visualize and infer that the rotations by Euler angles: $(\alpha,\beta,\gamma) =(\phi_i,\theta_i,0)$ in the $z-y1-z2$ sense, will align the local cartesian system at the pole with that defined at position $\hat{n}_i$. The Euler angles in the $z-y-z$ convention are related to those in the $z-y1-z2$ convection by the following rule: $(\alpha,\beta,\gamma)_{z-y-z} =(\gamma,\beta,\alpha)_{z-y1-z2}$.
